Detecting Manual Alterations in Biological Image Data Using Contrastive Learning and Pairwise Image Comparison

### 2025-05-22 slides/final/main.tex
final presentation

\begin{frame}{Goal of research}
\begin{block}{Ensure biological image integrity}
Develop a contrastive learning model for pairwise image comparison to:
\begin{itemize}
    \item Detect alterations (color jittering, crop, rotation, noise) 
    \item Select pairs of images with the same content
    \item Outperform existing state-of-the-art  models (Barlow Twins\footnote{{\tiny \textit{J. Zbontar et al.} Barlow Twins: Self-Supervised Learning via Redundancy Reduction // ICML, 2021.}}, SimCLR\footnote{{\tiny \textit{T. Chen et al.} A Simple Framework for Contrastive Learning of Visual Representations // ICML, 2021.}}) on cell datasets
\end{itemize}
\end{block}
\begin{center}
    \includegraphics[width=1\textwidth]{fig/alterations.png}
\end{center}
\end{frame}

@@

\begin{frame}{One-slide talk}
\begin{block}{The problem}
Detection of similar images despite modifications.
\end{block}

\medskip

\begin{columns}
    \begin{column}{0.42\textwidth}
        \includegraphics[width=0.95\textwidth]{fig/model-eval.png}
    \end{column}
    \begin{column}{0.58\textwidth}
        The model should process two images and output a value from [0, 1] --
        the likelihood that they are identical, up to modifications.

@@

\begin{frame}{Literature}
\begin{block}{Key Articles}
    \begin{itemize}
        \item \textbf{SimCLR}: Chen T. et al. "A Simple Framework for Contrastive Learning of Visual Representations", ICML 2020
        \item \textbf{Barlow Twins}: Zbontar J. et al. "Barlow Twins: Self-Supervised Learning via Redundancy Reduction", ICML 2021  
        \item \textbf{CLIP}: Radford A. et al. "Learning Transferable Visual Models From Natural Language Supervision", ICML 2021
        \item \textbf{Siamese Networks}: Melekhov I. et al. "Siamese Network Features for Image Matching", ICPR 2016
    \end{itemize}
\end{block}
\end{frame}

@@

\begin{frame}{Problem statement}
\begin{block}{Given biological image dataset}
    \begin{equation*}
        \mathcal{D} = \{d_i \in \mathcal{S},\ i \in [0, N)\},\quad 
        \mathcal{S} \subseteq \mathbb{R}^{H \times W \times C}
    \end{equation*}
\end{block}

\begin{block}{Pairwise similarity classification}
    For any $(x, y) \in \mathcal{S} \times \mathcal{S}$, learn mapping:
    \begin{equation*}
        \mathcal{M}: (x, y) \mapsto s \in [0, 1]
    \end{equation*}
    where:
    \begin{itemize}
        \item $s=1$: \textit{similar} pair (same content pre-alteration)
        \item $s=0$: \textit{dissimilar} pair (different content)
    \end{itemize}
\end{block}
\end{frame}

@@

\begin{frame}{Problem statement}
\begin{block}{Model decomposition}
    $$ \mathcal{M}(x, y) = h(f(x), f(y)) $$
    where:
    $$ f : \mathcal{S} \rightarrow \mathbb{R}^d \, \text{(encoder)} $$
    $$ h : \mathbb{R}^d \times \mathbb{R}^d \rightarrow [0, 1] \, \text{(classifier)} $$
\end{block}

\begin{block}{Success criterion}
    Maximize accuracy over pairwise comparisons:
    \begin{equation*}
    \text{Acc} = \frac{1}{|\mathcal{P}|} \sum_{(x,y) \in \mathcal{P}} \mathbb{I}\big(\mathcal{M}(x,y) = I(x,y)\big)
    \end{equation*}
    where $\mathcal{P}$ is test pairs, $I(x,y)$ ground truth similarity.
\end{block}
\end{frame}

@@

\begin{frame}{Solution}
\begin{columns}
    \column{0.53\textwidth}
    \begin{block}{Barlow Twins Adaptation}
        \vspace{3mm}
        Architecture:
        \begin{itemize}
            \item ResNet-50 backbone
            \item Projector
            \item Similarity head
        \end{itemize}
        Training specific:
        \begin{itemize}
            \item Parallel image augmentation
            \item AdamW optimizer with decreasing learning rate
            \item Performed on a specially selected dataset
        \end{itemize}

        \vspace{2mm}
        \textbf{Key Innovation:} \\
        Custom model's head and dataset \\
    \end{block}

@@

\begin{frame}{Computational experiment}
\begin{block}{Experimental Setup}
\begin{itemize}
    \item \textbf{Dataset}: 630 biological scans (animal and plant cells)
    \item \textbf{Train/Test Split}: 80\%/20\%
    \item \textbf{Training}: 100 epochs, AdamW optimizer ($\gamma_{start}=3\cdot10^{-3}, \gamma_{end}=5\cdot10^{-4}$)
\end{itemize}
\end{block}

\begin{block}{Evaluation Protocol}
\begin{itemize}
    \item Compare with Barlow Twins baseline
    \item Metrics:
    \begin{itemize}
        \item Accuracy 
        \item F1-Score, Precision, Recall
        \item AUC-ROC
    \end{itemize}
\end{itemize}
\end{block}
\end{frame}

@@

\begin{frame}{Computational Experiment}
\begin{columns}[T]
    \column{0.50\textwidth}
    \vspace{-4mm}
    \begin{block}{ROC-AUC Comparison}
        \centering
        \vspace{-2mm}
        {\tiny Model} \\
        \includegraphics[width=0.9\textwidth]{fig/roc-curve-model.png} \\
        \vspace{-2mm}
        {\tiny Barlow Twins} \\
        \includegraphics[width=0.9\textwidth]{fig/roc-curve-BT.png}
    \end{block}

    \column{0.50\textwidth}
    \vspace{-4mm}
    \begin{block}{Performance Metrics}
        \vspace{3mm}
        \centering
        \begin{tabular}{lcc}
        \hline
        Metric & Model & Barlow Twins \\ \hline
        Accuracy & \color{red}0.85 & 0.68 \\
        F1-Score & \color{red}0.80 & 0.48 \\
        Precision & \color{red}0.82 & 0.54 \\
        Recall & \color{red}0.78 & 0.43 \\ 
        AUC & \color{red}0.92 & 0.69 \\ \hline
        \end{tabular}
    \end{block}

    \begin{itemize}
        \footnotesize
        \item All metrics computed on test set (20\% data)
        \item Threshold = 0.5 for binary classification
    \end{itemize}
\end{columns}
\end{frame}

@@

\begin{frame}{Conclusion}
\begin{block}{Key achievements}
    \begin{itemize}
        \item Significant accuracy metrics improvement over state-of-the-art model
        \item Robust to 4 types of manual alterations
        \item First biological-SSL solution for:
        \begin{itemize}
            \item Automated fraud detection
            \item Image provenance verification
        \end{itemize}
    \end{itemize}
\end{block}
\end{frame}

>>>>

\begin{frame}{Ensure Biological Image Integrity}
\begin{block}{Comparing 2 images}
Develop a contrastive learning model for pairwise image comparison to:
\begin{enumerate}
    \item Detect alterations (color jittering, crop, rotation, noise) 
    \item Select pairs of images with the same content
    \item Outperform existing state-of-the-art  models (Barlow Twins\footnote{{\tiny \textit{J. Zbontar et al.} Barlow Twins: Self-Supervised Learning via Redundancy Reduction // ICML, 2021.}}, SimCLR\footnote{{\tiny \textit{T. Chen et al.} A Simple Framework for Contrastive Learning of Visual Representations // ICML, 2021.}}) on cell datasets
\end{enumerate}
\end{block}
\begin{center}
    \includegraphics[width=1\textwidth]{fig/alterations.png}
\end{center}
\end{frame}

@@

\begin{frame}{Detection of Similar Images Despite Modifications}

\begin{columns}
    \begin{column}{0.42\textwidth}
        \includegraphics[width=1\textwidth]{fig/model-example.png}
    \end{column}
    \begin{column}{0.58\textwidth}
        The model should process two images and output a value from [0, 1] --
        the likelihood that they are identical, up to modifications.

@@

\begin{frame}{Key Articles}
\begin{enumerate}
    \item \textbf{SimCLR}: Chen T. et al. "A Simple Framework for Contrastive Learning of Visual Representations", ICML 2020
    \item \textbf{Barlow Twins}: Zbontar J. et al. "Barlow Twins: Self-Supervised Learning via Redundancy Reduction", ICML 2021  
    \item \textbf{CLIP}: Radford A. et al. "Learning Transferable Visual Models From Natural Language Supervision", ICML 2021
    \item \textbf{Siamese Networks}: Melekhov I. et al. "Siamese Network Features for Image Matching", ICPR 2016
\end{enumerate}
\end{frame}

@@

\begin{frame}{Problem Statement}
\begin{block}{Given biological image dataset}
    \begin{equation*}
        \mathcal{D} = \{d_i \in \mathcal{S},\ i \in [0, N)\},\quad 
        \mathcal{S} \subseteq \mathbb{R}^{H \times W \times C}
    \end{equation*}
\end{block}

\begin{block}{Pairwise similarity classification}
    For any $(x, y) \in \mathcal{S} \times \mathcal{S}$, learn mapping:
    \begin{equation*}
        \mathcal{M}: (x, y) \mapsto s \in [0, 1]
    \end{equation*}
    where:
    $$
    s = 
    \begin{cases}
    1, & \text{\textit{similar} pair (same content pre-alteration)}, \\
    0, & \text{\textit{dissimilar} pair (different content)}
    \end{cases}
    $$
\end{block}
\end{frame}

@@

\begin{frame}{Model Decomposition}
\begin{block}{Fixed structure}
    $$ \mathcal{M}(x, y) = h(f(x), f(y)) $$
    where:
    $$ f : \mathcal{S} \rightarrow \mathbb{R}^d \, \text{(encoder)} $$
    $$ h : \mathbb{R}^d \times \mathbb{R}^d \rightarrow [0, 1] \, \text{(classifier)} $$
\end{block}

\begin{block}{Quality metric}
    Maximize accuracy over pairwise comparisons:
    \begin{equation*}
    \text{Acc} = \frac{1}{|\mathcal{P}|} \sum_{(x,y) \in \mathcal{P}} \mathbb{I}\big(\mathcal{M}(x,y) = I(x,y)\big)
    \end{equation*}
    where $\mathcal{P}$ is test pairs, $I(x,y)$ ground truth similarity.
\end{block}
\end{frame}

@@

\begin{frame}{Barlow Twins Adaptation (BTA)}
\begin{columns}
    \column{0.53\textwidth}
    \begin{block}{Model based on Barlow Twins}
        \vspace{3mm}
        Architecture:
        \begin{itemize}
            \item ResNet-50 backbone
            \item Projector
            \item Similarity head
        \end{itemize}
        Training specific:
        \begin{itemize}
            \item Parallel image augmentation
            \item AdamW optimizer with decreasing learning rate
            \item Performed on a specially selected dataset
        \end{itemize}

        \vspace{2mm}
        \textbf{Key Innovation:} \\
        Custom model's head, training dataset and pipeline \\
    \end{block}

@@

\begin{frame}{Comparing Training Pipelines}
\begin{block}{Modifying original learning specifics}
    \begin{columns}
        \column{0.6\textwidth}
        \vspace{2mm}
        \begin{center}
            \textbf{Consecutively}
        \end{center}
        $ \mathcal{L}_{proj} = \sum_i(1 - \mathcal{C}_{ii})^2 + \lambda_{proj} \sum_i \sum_{j \neq i} \mathcal{C}_{ij}^2 $

        \vspace{3mm}

        $ \mathcal{C}_{ij} = \frac{\sum_b{z_{b,i}^A z_{b,j}^B}}{\sqrt{\sum_b{(z_{b,i}^A)^2}} \sqrt{\sum_b{(z_{b,j}^B)^2}}} $

        \vspace{3mm}

        $ \mathcal{L}_{sim} = BCELoss $

        \column{0.4\textwidth}
        \vspace{2mm}
        \begin{center}
            \textbf{Parallel}
        \end{center}
        $ \mathcal{L} = \mathcal{L}_{sim} + \lambda \cdot \mathcal{L}_{proj} $
    \end{columns}
\end{block}
\end{frame}

@@

\begin{frame}{Experiment Setup}
    \textbf{Dataset}: 630 biological scans (animal and plant cells) \\
    \textbf{Train/Test Split}: 80\%/20\% \\
    \textbf{Training}: 100 epochs, AdamW optimizer ($\gamma_{start}=3\cdot10^{-3}, \gamma_{end}=5\cdot10^{-4}$)

\vspace{4mm}

\begin{block}{Evaluation Protocol}
    Compare with Barlow Twins and SimCLR baselines \\
    Metrics:
    \begin{enumerate}
        \item Accuracy 
        \item F1-Score, Precision, Recall
        \item AUC-ROC
    \end{enumerate}
\end{block}
\end{frame}

@@

\begin{frame}{Significant Accuracy Improvements}
\vspace{3mm}
\centering
\begin{tabular}{lcccc}
\hline
Metric & BTA conseq & BTA parallel & BT & SimCLR \\ \hline
Accuracy & 0.89 & \textbf{\color{red}0.90} & 0.68 & \textbf{0.67} \\
F1-Score & \textbf{\color{red}0.85} & 0.80 & \textbf{0.48} & 0.54 \\
Recall & \textbf{\color{red}0.84} & 0.71 & \textbf{0.43} & 0.52 \\
Precision & 0.87 & \textbf{\color{red}0.92} & 0.54 & \textbf{0.46} \\
AUC & 0.95 & \textbf{\color{red}0.97} & \textbf{0.69} & 0.71 \\ \hline
\end{tabular}

\vspace{6mm}
\begin{columns}[T]
    \column{0.5\textwidth}
    \centering
    {\tiny BTA conseq} \\
    \includegraphics[width=1\textwidth]{fig/roc-auc-BTA.png} \\

    \hspace{-8mm}

    \column{0.5\textwidth}
    \centering
    {\tiny BTA parallel} \\
    \includegraphics[width=1\textwidth]{fig/roc-auc-BTA-combined.png}

\end{columns}
\end{frame}

\begin{frame}{Key Achievements}
\begin{enumerate}
    \item Significant accuracy metrics improvement over state-of-the-art models Barlow Twins and SimCLR
    \item Model is robust to 4 types of manual alterations
    \item First biological-SSL solution for automated fraud detection and image provenance verification
\end{enumerate}

\begin{block}{Research materials are available in GitHub repository}
    \url{https://github.com/intsystems/2025-Project-170}
\end{block}
\end{frame}
